### 2016-12-14 paper/5_evaluation.tex
Refactor format of paper and some graphics

\textbf{Experimental Setup}. For our evaluation, we use a Google compute engine cluster with $8$ machines of type \code{n1-standard-16}.
Each machine is equipped with $60$ gigabytes of main memory and has $16$ virtual CPUs. 
The Spark execution engine runs on Apache Spark-2.0.0~\cite{spark}.
For the Flink execution engine, we use Apache Flink-1.1.2~\cite{flink}. 
As underlying distributed file system both systems use Apache Hadoop-2.7.3~\cite{hadoop:2008a}.

@@

\textbf{Matrix Multiplication}. 
As a first benchmark, we choose the matrix multiplication $A\times B$ with $A,B \in \mathbb{R}^{n\times n}$ and $n$ being the dimensionality. 
The matrix multiplication operation is demanding, both in CPU load as well as network I/O as it requires two network shuffles. 
The matrices $A$ and $B$ are sparse matrices with uniformly distributed non-zero cells. 
They are randomly generated prior to the matrix multiplication with a sparsity of $0.001$. 
As baseline, we run the same matrix multiplication on a single machine of the cluster using Gilbert's local execution engine. 
The Flink and Spark execution engines are both started with 48 gigabytes of main memory for their task managers and they distribute the work equally among all worker nodes. 
In the first experiment, we fixed the block sizes to $500 \times 500$ and set the number of cores to $64$. 
We then increased the dimensionality $n$ of $A$ and $B$ to observe the runtime behavior. 
The resulting execution times for the local, Flink and Spark execution engines are shown in \cref{fig:mmLoadRuntime}. 
In the second experiment, we investigate the scaling behavior of Gilbert with respect to the cluster size. 
In order to observe the communication costs, we keep the work load per core constant while increasing the number of cores. 
For this experiment we vary the number of cores from $1$ to $128$ and scale $n$ such that $n^3/\#cores$ is constant. 
We started with $n=5000$ for a single core and reached $n=25000$ on $128$ cores. 
The results of the experiment are shown in \cref{fig:mmNodesRuntime}.

@@

\begin{figure}[htbp]
	\centering
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{loglogaxis}[
				xlabel={Dimensionality $n$},
				ylabel={Execution time $t$ in s},
				width=\dualpgfwidth,
				height=.78\dualpgfwidth,
				legend entries={Local, Spark, Flink},
				legend pos=north west,
				font=\footnotesize
			]
			\addplot[
				color=blue,
				mark=x,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMult/matrixMultLoadReference};
			\addplot[
				color=red,
				mark=o,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMult/matrixMultLoadSpark};
			\addplot[
				color=teal,
				mark=triangle,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMult/matrixMultLoadFlink};
			\end{loglogaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:mmLoadRuntime}
	\end{subfigure}
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Number of cores},
				ylabel={Execution time $t$ in s},
				legend pos=north west,
				legend entries={Spark, Flink},
				ymin=0.0,
				ymax=2000,
				width=\dualpgfwidth,
				height=.78\dualpgfwidth,
				font=\footnotesize
			]
			\addplot[
				color=red,
				mark=o,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/matrixMult/matrixMultScalabilitySpark};
			\addplot[
				color=teal,
				mark=triangle,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/matrixMult/matrixMultScalabilityFlink};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:mmNodesRuntime}
	\end{subfigure}
	\caption{Scalability of matrix multiplication. \subref{fig:mmLoadRuntime} Execution time of matrix multiplication depending on the data size. \subref{fig:mmNodesRuntime} Execution time of matrix multiplication depending on the cluster size with constant work load per core.}
	\label{fig:mmScalability}
\end{figure}

\textbf{Gaussian Non-negative Matrix Factorization}. As second benchmark, we choose the Gaussian non-negative matrix factorization (GNMF) algorithm~\cite{seung:anips2001a}.
GNMF finds for a given matrix $V \in \mathbb{R}^{d\times w}$ a factorization $W \in \mathbb{R}^{d\times t}$ and $H \in \mathbb{R}^{t\times w}$ such that $V\approx W H$ holds true.
For our evaluation, we calculate one step of the GNMF algorithm.
We set $t=10$, $w=100000$ and vary the number of rows $d$ of $V$.
The matrix $V\in\mathbb{R}^{d\times 100000}$ is a sparse matrix whose sparsity is set to $0.001$.
The non-zero cells of $V$ are uniformly distributed.
As baseline, we run the GNMF on a single machine of the cluster using the local execution engine.
Like for the matrix multiplication benchmark, the task manager of Spark and Flink are started with 48 gigabytes of memory.
In the first experiment we fix the number of cores to $64$.
We start with $d=500$ and increase the number of rows of $V$ to $150000$.
The block size of Gilbert is set to $500 \times 500$.
In order to compare the results with the optimized GNMF MapReduce implementation proposed in~\cite{liu:2010a}, we re-implemented the algorithm using  Spark and Flink.
This hand-tuned implementation, henceforth denoted as HT-GNMF (HT in the graphs), is also executed on $64$ cores.
The execution times of HT-GNMF and Gilbert's GNMF are shown in \cref{fig:nmfLoadRuntime}.
In the second experiment of the GNMF benchmark, we analyze how Gilbert scales-out when increasing the cluster size while keeping the work load for each core constant.
We vary the cluster size from $1$ core to $64$ cores and scale the number of documents $d$ accordingly.
Initially we start with $1000$ documents and, consequently, calculate the matrix factorization for $64000$ documents on $64$ cores.
The results of this experiment are shown in \cref{fig:nmfNodesRuntime}.

\begin{figure}[htbp]
	\centering
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{loglogaxis}[
				xlabel={Number of rows of matrix $V$},
				ylabel={Execution time $t$ in s},
				width=\dualpgfwidth,
				height=.78\dualpgfwidth,
				legend pos=north west,
				legend columns=2,
				legend entries={Local, Spark, Flink, HT Spark, HT Flink},
				ymax=800,
				font=\footnotesize
			]
			\addplot[
				color=blue,
				mark=x,
			] table[
				x=Rows,
				y=Time,
			]
			{data/nnmfStep/nnmfStepLoadReference};

@@

			\addplot[
				color=teal,
				mark=triangle,
			]table[
				x=Rows,
				y=Time,
			]
			{data/nnmfStep/nnmfStepLoadFlink};
			\addplot[
				color=black,
				mark=diamond,
			]table[
				x=rowsV,
				y=time,
			]
			{data/nnmfStep/gnmfStepBenchLoadSpark};
			\addplot[
				color=magenta,
				mark=square,
			]table[
				x=rowsV,
				y=time,
			]
			{data/nnmfStep/gnmfStepBenchLoadFlink};
			\end{loglogaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:nmfLoadRuntime}
	\end{subfigure}
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Number of rows of matrix $V$},
				ylabel={Speedup of HT over Gilbert},
				width=\dualpgfwidth,
				height=.78\dualpgfwidth,
				ymin=0,
				ymax=2.5,
				legend entries={Spark,Flink},
				legend pos=north west,
				font=\footnotesize
			]
			\addplot[
				color=red,
				mark=o,
			] table[
				x=Rows,
				y=Speedup,
			]
			{data/nnmfStep/gnmfStepSpeedupSpark};

			\addplot[
				color=teal,
				mark=diamond,
			] table[
				x=Rows,
				y=Speedup,
			]
			{data/nnmfStep/gnmfStepSpeedupFlink};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:gnmfSpeedup}
	\end{subfigure}
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Number of cores},
				ylabel={Execution time $t$ in s},
				xmax=101,
				ymin=0,
				ymax=60,
				legend entries={Spark, Flink},
				legend pos=north west,
				width=\dualpgfwidth,
				height=.7\dualpgfwidth,
				font=\footnotesize
			]
			\addplot[
				color=red,
				mark=o,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/nnmfStep/nnmfStepScalabilitySpark};

			\addplot[
				color=teal,
				mark=diamond,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/nnmfStep/nnmfStepScalabilityFlink};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:nmfNodesRuntime}
	\end{subfigure}
	\caption{Scalability of GNMF. \subref{fig:nmfLoadRuntime} Execution time of one GNMF step depending on the data size. \subref{fig:gnmfSpeedup} Speedup of HT-GNMF over Gilbert's GNMF using Spark and Flink. \subref{fig:nmfNodesRuntime} Execution time of one GNMF step depending on the cluster size with constant work load per core.}
	\label{fig:nmfBenchmark}
\end{figure}

@@

In this section, we want to investigate how well the PageRank algorithm~\cite{page:1999a} implemented in Gilbert performs compared to a specialized implementation (denoted as SP in the figures).
We expect that the Gilbert runtime adds some overhead.
Furthermore, the high-level linear algebra abstraction of Gilbert might make it difficult to exploit certain properties to speed up the processing.
We first execute PageRank directly in Gilbert, given the Gilbert code in \cref{lst:gilbertPageRank}, and then run them directly on Flink and Spark.
In contrast to Gilbert, the direct implementation requires a deep understanding of the underlying runtime system.
Furthermore, the distributed implementation is far from trivial compared to the linear algebra representation of the original problem.
\begin{listing}[htbp]
	\begin{lstlisting}[language=Matlab,
		commentstyle=\color{black},
		  stringstyle=\color{black},
		  keywordstyle=\color{black}\bfseries,
		  morekeywords={ones, fixpoint},
		  xleftmargin=.0\textwidth,
		  basicstyle=\footnotesize]
% load adjacency matrix
A = load();
maxIterations = 10;
d = sum(A, 2); % outdegree per vertex
% create the column-stochastic transition matrix
T = (diag(1 ./ d) * A)'; 
r_0 = ones(numVertices, 1) / numVertices; % initialize the ranks
e = ones(numVertices, 1) / numVertices;
% PageRank calculation
fixpoint(r_0, @(r) .85 * T * r + .15 * e, maxIterations)
	\end{lstlisting}
	\caption{Gilbert PageRank implementation.}
	\label{lst:gilbertPageRank}
\end{listing}
The specialized Spark and Flink implementation of PageRank works as follows.
The Page\-Rank vector is represented by a set of tuples $(w_i, r_i)$ with $w_i$ denoting the web site $i$ and $r_i$ being its rank.
The adjacency matrix is stored row-wise as a set of tuples $(w_i, A_i)$ with $w_i$ denoting the web site $i$ and $A_i$ being its adjacency list.
For the adjacency list $A_i$, it holds that $w_j \in A_i$ if and only if there exists a link from web site $i$ to $j$.
The next PageRank vector is calculated as follows.
The PageRank $r_i$ of web site $i$ is joined with its adjacency list $A_i$.
For each outgoing link $w_j \in A_i$ a new tuple $(w_j, 0.85r_i/\left|A_i\right|)$ with the rank of $i$ being divided by the number of outgoing links is created.
In order to incorporate the random jump behavior to any available web site, a tuple $(w_i, 0.15/|w|)$, with $|w|$ being the number of all web sites, is generated for each web site $i$.
In order to compute the next PageRank vector, all newly generated tuples are grouped according to their ID and summed up.
These steps constitute a single PageRank iteration whose data flow plan is depicted in \cref{fig:pageRankDataFlow}.
\begin{figure}[htbp]
	\centering
	\includegraphics[width=.275\linewidth]{images/pageRankStep.png}
	\caption{Data flow of one iteration of the PageRank algorithm for Spark and Flink.}
	\label{fig:pageRankDataFlow}
\end{figure}
\textbf{Experiment}. For comparison, $10$ steps of the PageRank algorithm for varying sizes of the adjacency matrix $A$ are calculated.
The randomly generated adjacency matrix $A$ is a sparse matrix of size $n \times n$ with a sparsity of $0.001$.
The computation is executed on $64$ cores with a block size of $500 \times 500$.
The execution times are depicted in \cref{fig:pageRankResults}.
\begin{figure}[htbp]
	\centering
	\begin{subfigure}[h]{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{loglogaxis}[
				xlabel={Number of vertices $n$},
				ylabel={Execution time $t$ in s},
				xmax=250000,
				ymax=50000,
				legend pos=north west,
				legend entries={Spark, Flink, SP Flink, SP Spark},
				width=\dualpgfwidth,
				height=.78\dualpgfwidth,
				font=\footnotesize
			]

@@

			\addplot[black,
				mark=triangle,
			] table[
				x=rows,
				y=time,
			]
			{data/pagerank/pagerankBenchLoadSpark};
			\end{loglogaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:pageRankResults}
	\end{subfigure}
	\begin{subfigure}[h]{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Number of vertices $n$},
				ylabel={Speedup},
				legend pos=north west,
				legend entries={SP Spark, SP Flink},
				width=\dualpgfwidth,
				height=.78\dualpgfwidth,
				font=\footnotesize
			]

@@

			\addplot[red,
				mark=o,
			] table[
				x=NumVertices,
				y=Speedup,
			]
			{data/pagerank/pagerankSpeedupFlink};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:pageRankSpeedup}
	\end{subfigure}
	\caption{Comparison of Gilbert's PageRank implementation with specialized algorithms on Spark and Flink. \subref{fig:pageRankResults} Execution time of $10$ steps of the PageRank algorithm depending on the adjacency matrix's size. \subref{fig:pageRankSpeedup} Speedup of specialized algorithms with respect to Gilbert's implementations.}
	\label{fig:pageRankEvaluation}
\end{figure}
The graphs in \cref{fig:pageRankResults} show that the specialized PageRank algorithm runs faster than Gilbert's versions of PageRank.
Furthermore, the specialized algorithms show a better scalability.
The speedup of the specialized algorithms with respect to Gilbert's implementations for this experiment is shown in \cref{fig:pageRankSpeedup}.
For $n\le 50000$, the hand-tuned algorithms running on Spark and Flink show a similar speedup.
The Flink and Spark version achieve a speedup of approximately $13$ and $12$ for $n=50000$, respectively.
However, for $n = 100000$ we can observe a speedup of $76$ for the Flink specialized implementation whereas Spark reaches a speedup of $14$.
Gilbert's performance degradation with Flink is likley caused by earlier data spilling.

>>>>

\textbf{Experimental Setup}. For our evaluation, we use a Google compute engine cluster with $8$ machines of type \code{n1-standard-16}.
Each machine is equipped with $60$ gigabytes of main memory and has $16$ virtual CPUs. 
The Spark execution engine runs on Apache Spark-2.0.0~\cite{spark}.
For the Flink execution engine, we use Apache Flink-1.1.2~\cite{flink}. 
As underlying distributed file system we use Apache Hadoop-2.7.3~\cite{hadoop:2008a}.

@@

\textbf{Matrix Multiplication}. 
As a first benchmark, we choose the matrix multiplication $A\times B$ with $A,B \in \mathbb{R}^{n\times n}$ and $n$ being the dimensionality. 
The matrix multiplication operation is demanding, both in CPU load as well as network I/O as it requires two network shuffles. 
The matrices $A$ and $B$ are sparse matrices with uniformly distributed non-zero cells. 
They are randomly generated prior to the matrix multiplication with a sparsity of $0.001$. 
As baseline, we run the same matrix multiplication on a single machine of the cluster using Gilbert's local execution engine. 
The Flink and Spark execution engines are both started with $48$ gigabytes of main memory for their task managers and they distribute the work equally among all worker nodes. 
In the first experiment, we fixed the block sizes to $500 \times 500$ and set the number of cores to $64$. 
We then increased the dimensionality $n$ of $A$ and $B$ to observe the runtime behavior. 
The resulting execution times for the local, Flink and Spark execution engines are shown in \cref{fig:mmLoadRuntime}. 
In the second experiment, we investigate the scaling behavior of Gilbert with respect to the cluster size. 
In order to observe the communication costs, we keep the work load per core constant while increasing the number of cores. 
For this experiment we vary the number of cores from $1$ to $128$ and scale $n$ such that $n^3/\#cores$ is constant. 
We started with $n=5000$ for a single core and reached $n=25000$ on $128$ cores. 
The results of the experiment are shown in \cref{fig:mmNodesRuntime}.
\begin{figure}[htbp]
	\centering
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{loglogaxis}[
				xlabel={Dimensionality $n$},
				ylabel={Execution time $t$ in s},
				width=\dualpgfwidth,
				height=.78\dualpgfwidth,
				legend entries={Local, Spark, Flink},
				legend pos=north west,
				legend style={font=\scriptsize},
				font=\footnotesize
			]
			\addplot[
				color=blue,
				mark=x,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMult/matrixMultLoadReference};
			\addplot[
				color=red,
				mark=o,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMult/matrixMultLoadSpark};
			\addplot[
				color=teal,
				mark=triangle,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMult/matrixMultLoadFlink};
			\end{loglogaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:mmLoadRuntime}
	\end{subfigure}
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Number of cores},
				ylabel={Execution time $t$ in s},
				legend pos=north west,
				legend entries={Spark, Flink},
				legend style={font=\scriptsize},
				ymin=0.0,
				ymax=2000,
				width=\dualpgfwidth,
				height=.78\dualpgfwidth,
				font=\footnotesize
			]
			\addplot[
				color=red,
				mark=o,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/matrixMult/matrixMultScalabilitySpark};
			\addplot[
				color=teal,
				mark=triangle,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/matrixMult/matrixMultScalabilityFlink};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:mmNodesRuntime}
	\end{subfigure}
	\caption{Scalability of matrix multiplication. \subref{fig:mmLoadRuntime} Execution time of matrix multiplication depending on the data size. \subref{fig:mmNodesRuntime} Execution time of matrix multiplication depending on the cluster size with constant work load per core.}
	\label{fig:mmScalability}
\end{figure}

@@

\textbf{Gaussian Non-negative Matrix Factorization}. As second benchmark, we choose the Gaussian non-negative matrix factorization (GNMF) algorithm~\cite{seung:anips2001a}.
GNMF finds for a given matrix $V \in \mathbb{R}^{d\times w}$ a factorization $W \in \mathbb{R}^{d\times t}$ and $H \in \mathbb{R}^{t\times w}$ such that $V\approx W H$ holds true.
For our evaluation, we calculate one step of the GNMF algorithm.
We set $t=10$, $w=100000$ and vary the number of rows $d$ of $V$.
The matrix $V\in\mathbb{R}^{d\times 100000}$ is a sparse matrix whose sparsity is set to $0.001$.
The non-zero cells of $V$ are uniformly distributed.
As baseline, we run the GNMF on a single machine of the cluster using the local execution engine.
Like for the matrix multiplication benchmark, the task manager of Spark and Flink are started with $48$ gigabytes of memory.
In the first experiment we fix the number of cores to $64$.
We start with $d=500$ and increase the number of rows of $V$ to $150000$.
The block size of Gilbert is set to $500 \times 500$.
In order to compare the results with the optimized GNMF MapReduce implementation proposed in~\cite{liu:2010a}, we re-implemented the algorithm using  Spark and Flink.
This hand-tuned implementation, henceforth denoted as HT-GNMF (HT in the graphs), is also executed on $64$ cores.
The execution times of HT-GNMF and Gilbert's GNMF are shown in \cref{fig:nmfLoadRuntime}.
In the second experiment of the GNMF benchmark, we analyze how Gilbert scales-out when increasing the cluster size while keeping the work load for each core constant.
We vary the cluster size from $1$ core to $64$ cores and scale the number of documents $d$ accordingly.
Initially we start with $1000$ documents and, consequently, calculate the matrix factorization for $64000$ documents on $64$ cores.
The results of this experiment are shown in \cref{fig:nmfNodesRuntime}.
\begin{figure}[!h]
	\centering
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{loglogaxis}[
				xlabel={Number of rows of matrix $V$},
				ylabel={Execution time $t$ in s},
				width=\dualpgfwidth,
				height=.16\paperheight,
				legend pos=north west,
				legend columns=2,
				legend entries={Local, Spark, Flink, HT Spark, HT Flink},
				legend style={font=\scriptsize},
				ymax=800,
				font=\footnotesize
			]
			\addplot[
				color=blue,
				mark=x,
			] table[
				x=Rows,
				y=Time,
			]
			{data/nnmfStep/nnmfStepLoadReference};

@@

			\addplot[
				color=teal,
				mark=triangle,
			]table[
				x=Rows,
				y=Time,
			]
			{data/nnmfStep/nnmfStepLoadFlink};
			\addplot[
				color=black,
				mark=diamond,
			]table[
				x=rowsV,
				y=time,
			]
			{data/nnmfStep/gnmfStepBenchLoadSpark};
			\addplot[
				color=magenta,
				mark=square,
			]table[
				x=rowsV,
				y=time,
			]
			{data/nnmfStep/gnmfStepBenchLoadFlink};
			\end{loglogaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:nmfLoadRuntime}
	\end{subfigure}
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Number of rows of matrix $V$},
				ylabel={Speedup of HT over Gilbert},
				width=\dualpgfwidth,
				height=.16\paperheight,
				ymin=0,
				ymax=2.5,
				legend entries={Spark,Flink},
				legend pos=north west,
				legend style={font=\scriptsize},
				font=\footnotesize
			]
			\addplot[
				color=red,
				mark=o,
			] table[
				x=Rows,
				y=Speedup,
			]
			{data/nnmfStep/gnmfStepSpeedupSpark};

			\addplot[
				color=teal,
				mark=diamond,
			] table[
				x=Rows,
				y=Speedup,
			]
			{data/nnmfStep/gnmfStepSpeedupFlink};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:gnmfSpeedup}
	\end{subfigure}
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Number of cores},
				ylabel={Execution time $t$ in s},
				xmax=101,
				ymin=0,
				ymax=60,
				legend entries={Spark, Flink},
				legend pos=north west,
				legend style={font=\scriptsize},
				width=\dualpgfwidth,
				height=.14\paperheight,
				font=\footnotesize]
			\addplot[
				color=red,
				mark=o,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/nnmfStep/nnmfStepScalabilitySpark};
			\addplot[
				color=teal,
				mark=diamond,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/nnmfStep/nnmfStepScalabilityFlink};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:nmfNodesRuntime}
	\end{subfigure}
	\caption{Scalability of GNMF. \subref{fig:nmfLoadRuntime} Execution time of one GNMF step depending on the data size. \subref{fig:gnmfSpeedup} Speedup of HT-GNMF over Gilbert's GNMF using Spark and Flink. \subref{fig:nmfNodesRuntime} Execution time of one GNMF step depending on the cluster size with constant work load per core.}
	\label{fig:nmfBenchmark}
\end{figure}

@@

In this section, we want to investigate how well the PageRank algorithm~\cite{page:1999a} implemented in Gilbert performs compared to a specialized implementation (denoted as SP in the figures).
We expect that the Gilbert runtime adds some overhead.
Furthermore, the high-level linear algebra abstraction of Gilbert might make it difficult to exploit certain properties to speed up the processing.
We first execute PageRank directly in Gilbert, given the Gilbert code in \cref{lst:gilbertPageRank}, and then run them directly on Flink and Spark.
In contrast to Gilbert, the direct implementation requires a deep understanding of the underlying runtime system.
Furthermore, the distributed implementation is far from trivial compared to the linear algebra representation of the original problem.
\begin{listing}[htbp]
	\begin{lstlisting}[language=Matlab,
		commentstyle=\color{black},
		  stringstyle=\color{black},
		  keywordstyle=\color{black}\bfseries,
		  morekeywords={ones, fixpoint},
		  xleftmargin=.0\textwidth,
		  basicstyle=\small]
% load adjacency matrix
A = load();
maxIterations = 10;
d = sum(A, 2); % outdegree per vertex
% create the column-stochastic transition matrix
T = (diag(1 ./ d) * A)'; 
r_0 = ones(numVertices, 1) / numVertices; % initialize the ranks
e = ones(numVertices, 1) / numVertices;
% PageRank calculation
fixpoint(r_0, @(r) .85 * T * r + .15 * e, maxIterations)
	\end{lstlisting}
	\caption{Gilbert PageRank implementation.}
	\label{lst:gilbertPageRank}
\end{listing}

The specialized Spark and Flink implementation of PageRank works as follows.
The Page\-Rank vector is represented by a set of tuples $(w_i, r_i)$ with $w_i$ denoting the web site $i$ and $r_i$ being its rank.
The adjacency matrix is stored row-wise as a set of tuples $(w_i, A_i)$ with $w_i$ denoting the web site $i$ and $A_i$ being its adjacency list.
For the adjacency list $A_i$, it holds that $w_j \in A_i$ if and only if there exists a link from web site $i$ to $j$.
The next PageRank vector is calculated as follows.
The PageRank $r_i$ of web site $i$ is joined with its adjacency list $A_i$.
For each outgoing link $w_j \in A_i$ a new tuple $(w_j, 0.85r_i/\left|A_i\right|)$ with the rank of $i$ being divided by the number of outgoing links is created.
In order to incorporate the random jump behavior to any available web site, a tuple $(w_i, 0.15/|w|)$, with $|w|$ being the number of all web sites, is generated for each web site $i$.
In order to compute the next PageRank vector, all newly generated tuples are grouped according to their ID and summed up.
These steps constitute a single PageRank iteration whose data flow plan is depicted in \cref{fig:pageRankDataFlow}.
\begin{figure}[htbp]
	\centering
	\includegraphics[width=.275\linewidth]{images/pageRankStep.png}
	\caption{Data flow of one iteration of the PageRank algorithm for Spark and Flink.}
	\label{fig:pageRankDataFlow}
\end{figure}

\textbf{Experiment}. For comparison, $10$ steps of the PageRank algorithm for varying sizes of the adjacency matrix $A$ are calculated.
The randomly generated adjacency matrix $A$ is a sparse matrix of size $n \times n$ with a sparsity of $0.001$.
The computation is executed on $64$ cores with a block size of $500 \times 500$.
The execution times are depicted in \cref{fig:pageRankResults}.
\begin{figure}[htbp]
	\centering
	\begin{subfigure}[h]{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{loglogaxis}[
				xlabel={Number of vertices $n$},
				ylabel={Execution time $t$ in s},
				xmax=250000,
				ymax=50000,
				legend pos=north west,
				legend entries={Spark, Flink, SP Flink, SP Spark},
				legend style={font=\scriptsize},
				width=\dualpgfwidth,
				height=.78\dualpgfwidth,
				font=\footnotesize
			]

@@

			\addplot[black,
				mark=triangle,
			] table[
				x=rows,
				y=time,
			]
			{data/pagerank/pagerankBenchLoadSpark};
			\end{loglogaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:pageRankResults}
	\end{subfigure}
	\begin{subfigure}[h]{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Number of vertices $n$},
				ylabel={Speedup},
				legend pos=north west,
				legend entries={SP Spark, SP Flink},
				legend style={font=\scriptsize},
				width=\dualpgfwidth,
				height=.78\dualpgfwidth,
				font=\footnotesize
			]

@@

			\addplot[red,
				mark=o,
			] table[
				x=NumVertices,
				y=Speedup,
			]
			{data/pagerank/pagerankSpeedupFlink};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:pageRankSpeedup}
	\end{subfigure}
	\caption{Comparison of Gilbert's PageRank implementation with specialized algorithms on Spark and Flink. \subref{fig:pageRankResults} Execution time of $10$ steps of the PageRank algorithm depending on the adjacency matrix's size. \subref{fig:pageRankSpeedup} Speedup of specialized algorithms with respect to Gilbert's implementations.}
	\label{fig:pageRankEvaluation}
\end{figure}

The graphs in \cref{fig:pageRankResults} show that the specialized PageRank algorithm runs faster than Gilbert's versions of PageRank.
Furthermore, the specialized algorithms show a better scalability.
The speedup of the specialized algorithms with respect to Gilbert's implementations for this experiment is shown in \cref{fig:pageRankSpeedup}.
For $n\le 50000$, the hand-tuned algorithms running on Spark and Flink show a similar speedup.
The Flink and Spark version achieve a speedup of approximately $13$ and $12$ for $n=50000$, respectively.
However, for $n = 100000$ we can observe a speedup of $76$ for the Flink specialized implementation whereas Spark reaches a speedup of $14$.
Gilbert's performance degradation with Flink is likley caused by earlier data spilling.

### 2016-12-14 paper/6_related_work.tex
Refactor format of paper and some graphics

{\em Ricardo}~\cite{das:2010a} uses existing technologies to implement a scalable system for statistical analysis: it executes the data shipment via Hadoop and runs the actual analytic computation by R. 
It integrates these systems via Jaql~\cite{beyer:2011a}, a declarative high-level language for data-processing on Hadoop. 
Thereby, the user can initiate the distribution, transformation and aggregation of data within Hadoop. 
Furthermore, the system supports to run R code on the worker nodes as data transformations. 
However, in this setting the user still has to explicitly specify the data-flow and the data distribution, which requires a substantial understanding of MapReduce's principles. 
{\em RHIPE}~\cite{guha:s2012a} also tries to extend R to scale to large data sets using Hadoop. 
RHIPE follows an approach called divide and recombine. 
The idea is to split the examined data up into several chunks so that they fit in the memory of a single computer. 
Next a collection of analytic methods is applied to each chunk without communicating with any other computer. 
After each chunk is evaluated, the results will be recombined in an aggregation step to create the overall analytic result. 
However, this strict execution pipeline constrains the analysis process considerably. 
A system, which integrates more seamlessly into the R ecosystem, is {\em pR} (parallel R)~\cite{samatova:2009a}. 
The goal is to let statisticians compute large-scale data without having to learn a new system. 
pR achieves its goal by providing a set of specialized libraries which offer parallelized versions of different algorithms using MPI.
However, MPI does not integrate well in an evironment where clusters are shared and usually execute several jobs concurrenlty. 
Furthermore, it inherently lacks important properties such as fault-tolerance, elasticity and reliability. 
Another related system for R is {\em rmpi}~\cite{rmpi}. 
As this is only a wrapper for the respective MPI calls, it also suffers from the aforementioned problems. 
Furthermore, there is the SparkR~\cite{Venkataraman2016} project which aims to integrate Spark's API and SQL processing capabilities into R.

>>>>

{\em Ricardo}~\cite{das:2010a} uses existing technologies to implement a scalable system for statistical analysis: it executes the data shipment via Hadoop and runs the actual analytic computation by R. 
It integrates these systems via Jaql~\cite{beyer:2011a}, a declarative high-level language for data-processing on Hadoop. 
Thereby, the user can initiate the distribution, transformation and aggregation of data within Hadoop. 
Furthermore, the system supports to run R code on the worker nodes as data transformations. 
However, in this setting the user still has to explicitly specify the data-flow and the data distribution, which requires a substantial understanding of MapReduce's principles. 
{\em RHIPE}~\cite{guha:s2012a} also tries to extend R to scale to large data sets using Hadoop. 
RHIPE follows an approach called divide and recombine. 
The idea is to split the examined data up into several chunks so that they fit in the memory of a single computer. 
Next a collection of analytic methods is applied to each chunk without communicating with any other computer. 
After all chunks are evaluated, the results will be recombined in an aggregation step to create the overall analytic result. 
However, this strict execution pipeline constrains the analysis process considerably. 
A system which integrates more seamlessly into the R ecosystem is {\em pR} (parallel R)~\cite{samatova:2009a}. 
The goal is to let statisticians compute large-scale data without having to learn a new system. 
pR achieves its goal by providing a set of specialized libraries which offer parallelized versions of different algorithms using MPI.
However, MPI does not integrate well with an evironment where clusters are shared and usually execute several jobs concurrently. 
Furthermore, it lacks important properties such as fault-tolerance, elasticity and reliability. 
Another related system for R is {\em rmpi}~\cite{rmpi}. 
As this is only a wrapper for the respective MPI calls, it also suffers from the aforementioned problems. 
Furthermore, there is the SparkR~\cite{Venkataraman2016} project which aims to integrate Spark's API and SQL processing capabilities into R.
